---
title: 'Task 1: Principal Component Analysis'
author: "Nick McManus"
date: "2023-02-27"
output:
  rmdformats::downcute:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
---

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse) ## always
library(ggfortify) ## For PCA biplot
```

# Introduction:

In this task, we explore patterns within a multivariate food nutrition dataset using a principal component analysis (PCA). Nutrition data comes from the United States Department of Agriulture (USDA) and is organized into 25 food group categories, such as snacks, baby foods, beef products, etc. Variables for each food product include both macronutrients (i.e. fats, protein, carbohydrates) and micronutrients (i.e. vitamins and minerals). Looking at the concentration of vitamins and minerals in raw vegetables and fruits. 


Look at how they compare against snacks? 

**Source:** https://fdc.nal.usda.gov/index.html


# Read in and wrangle data:

The raw USDA nutrient data will be wrangled to return only information on the vitamins and minerals concentrations for raw fruits and vegetables.
```{r}
### read in data
usda <- read_csv('usda_nutrients.csv') %>% 
  janitor::clean_names() 

### explore the food group categories (type and number)
# unique(usda$food_group)

usda_types <- usda %>%
  group_by(food_group) %>%
  summarize(count = n())
```

```{r}
### create subset w/only raw veggies and fruits (no juice) 
### and info on vitamins and minerals


fruit_veg_micro <- usda %>% 
  ## only keep fruits and veggies
  filter(food_group %in% c("Vegetables and Vegetable Products",
                           "Fruits and Fruit Juices"),
         ## only keep raw, whole foods (no juice)
         str_detect(descrip, "raw"),
         !str_detect(descrip, "juice")) %>% 
  ## remove macronutrients and RDA ratios
  select(!ends_with(c("usrda", "_g", "_kcal"))) %>% 
  ## make food groups factors for later biplot
  mutate(food_group = factor(food_group, 
                             levels = c('Vegetables and Vegetable Products', 
                                        'Fruits and Fruit Juices'))) 

fruit_veg_macro <- usda %>% 
  ## only keep fruits and veggies
  filter(food_group %in% c("Vegetables and Vegetable Products",
                           "Fruits and Fruit Juices"),
         ## only keep raw, whole foods (no juice)
         str_detect(descrip, "raw"),
         !str_detect(descrip, "juice")) %>% 
  ## remove macronutrients and RDA ratios
  select(food_group, energy_kcal:fiber_g) %>%
  ## make food groups factors for later biplot
  mutate(food_group = factor(food_group, 
                             levels = c('Vegetables and Vegetable Products', 
                                        'Fruits and Fruit Juices')))

fruit_veg_all <- usda %>% 
  ## only keep fruits and veggies
  filter(food_group %in% c("Vegetables and Vegetable Products",
                           "Fruits and Fruit Juices"),
         ## only keep raw, whole foods (no juice)
         str_detect(descrip, "raw"),
         !str_detect(descrip, "juice")) %>% 
  ## remove macronutrients and RDA ratios
  select(!ends_with(c("usrda"))) %>%
  ## make food groups factors for later biplot
  mutate(food_group = factor(food_group, 
                             levels = c('Vegetables and Vegetable Products', 
                                        'Fruits and Fruit Juices')))
 
```



# Run PCA

``` {r}
### Create subset with only numeric values, then run PCA.
### Different units, so scale variables
fruit_veg_micro_pca <- fruit_veg_micro %>%
  select(where(is.numeric)) %>% 
  prcomp(scale = TRUE) 

fruit_veg_macro_pca <- fruit_veg_macro %>%
  select(where(is.numeric)) %>% 
  prcomp(scale = TRUE) 

fruit_veg_all_pca <- fruit_veg_all %>%
  select(where(is.numeric)) %>% 
  prcomp(scale = TRUE)

### See the loadings and summary info
# veggies_pca$rotation
# summary(veggies_pca)
```



# Visualize PCA

Create a biplot to look at PCA results in two-dimensions.
```{r}
autoplot(fruit_veg_macro_pca,
     	data = fruit_veg_macro,
     	loadings = TRUE,
     	colour = 'food_group',
     	alpha = .6,
     	loadings.label = TRUE,
     	loadings.colour = "black",
     	loadings.label.colour = "black",
     	loadings.label.vjust = -0.5) +
  scale_colour_manual(values = c('seagreen3', 'mediumorchid1'), 
                      labels = c('Vegetables', 'Fruit')) +
  theme_minimal() +
   theme(
    legend.position = c(.8, .8),
    axis.title.x = element_text(vjust = -1)
  ) +
  ## change legend dot size
  guides(colour = guide_legend(override.aes = list(size = 3), 
                               title = "Food Group:"))
 


autoplot(fruit_veg_micro_pca,
     	data = fruit_veg_micro,
     	loadings = TRUE,
     	colour = 'food_group',
     	loadings.label = TRUE,
     	loadings.colour = "black",
     	loadings.label.colour = "black",
     	loadings.label.vjust = -0.5
     	) +
  theme_minimal()

autoplot(fruit_veg_all_pca,
     	data = fruit_veg_micro,
     	loadings = TRUE,
     	colour = 'food_group',
     	loadings.label = TRUE,
     	loadings.colour = "black",
     	loadings.label.colour = "black",
     	loadings.label.vjust = -0.5
     	) +
  theme_minimal() 
```
**Figure 1.** Caption here. 