---
title: 'Task 1: Principal Component Analysis'
author: "Nick McManus"
date: "2023-02-27"
output:
  rmdformats::downcute:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
---

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse) ## always
library(ggfortify) ## For PCA biplot
```

# Introduction

In this task, we explore patterns within a multivariate food nutrition dataset using a principal component analysis (PCA). Nutrition data comes from the United States Department of Agriulture (USDA) National Nutrient Database and is organized into 25 food group categories, such as snacks, baby foods, beef products, etc. Variables for each food product include both macronutrients (i.e. fats, protein, carbohydrates) and micronutrients (i.e. vitamins and minerals). A PCA is first performed on the macronutrients of raw fruits and vegetables. This is then compared against a PCA conducted on all the nutrient information for raw fruits and vegetables.

**Source:** https://fdc.nal.usda.gov/index.html

# Read in and wrangle data

The raw USDA nutrient data will be wrangled to return only information on raw fruits and vegetables. Then, two data subsets are created; the first contains only the macronutrient information, while the second includes both macro and micronutrients.
```{r}
### read in data
usda <- read_csv('usda_nutrients.csv') %>% 
  janitor::clean_names() 

### explore the food group categories (type and number)
# unique(usda$food_group)

# usda_types <- usda %>%
#   group_by(food_group) %>%
#   summarize(count = n())
```

```{r}
### create subset w/only raw veggies and fruits (no juice) 
fruit_veg <- usda %>% 
  ## only keep fruits and veggies
  filter(food_group %in% c("Vegetables and Vegetable Products",
                           "Fruits and Fruit Juices"),
         ## only keep raw, whole foods (no juice)
         str_detect(descrip, "raw"),
         !str_detect(descrip, "juice"))


### Create subset with macronutrient info
fruit_veg_macro <- fruit_veg %>% 
  ## remove micronutrients and RDA ratios
  select(food_group, energy_kcal:fiber_g) %>%
  ## make food groups factors for later biplot
  mutate(food_group = factor(food_group, 
                             levels = c('Vegetables and Vegetable Products', 
                                        'Fruits and Fruit Juices')))


### Create subset with all info (except RDA)
fruit_veg_all <- fruit_veg %>% 
  ## remove RDA ratios (redundant info)
  select(!ends_with(c("usrda"))) %>%
  select(!"id") %>% 
  ## make food groups factors for later biplot
  mutate(food_group = factor(food_group, 
                             levels = c('Vegetables and Vegetable Products', 
                                        'Fruits and Fruit Juices')))
```



# Run PCAs
``` {r}
### Create subset with only numeric values, then run PCA.
### Different units, so scale variables

fruit_veg_macro_pca <- fruit_veg_macro %>%
  select(where(is.numeric)) %>% 
  prcomp(scale = TRUE) 

fruit_veg_all_pca <- fruit_veg_all %>%
  select(where(is.numeric)) %>% 
  prcomp(scale = TRUE)

### See the loadings and summary info
# fruit_veg_macro_pca$rotation
# summary(fruit_veg_macro_pca)
```



# Visualize PCAs

Two biplots are created to look at the PCA results in two-dimensions.
```{r}
### Macro biplot
autoplot(fruit_veg_macro_pca,
     	data = fruit_veg_macro,
     	loadings = TRUE,
     	colour = 'food_group',
     	size = 1.8,
     	alpha = .6,
     	loadings.label = TRUE,
     	loadings.colour = "black",
     	loadings.label.colour = "black",
     	loadings.label.vjust = -0.5) +
  ## dot colors and legend labels
  scale_colour_manual(values = c('seagreen3', 'mediumorchid1'), 
                      labels = c('Vegetables', 'Fruits')) +
  theme_minimal() +
  theme(
     ## legend  
     legend.position = c(.8, .8),
     legend.title = element_text(face = 'bold', size = 10),
     legend.box.background = element_rect(color = 'black', size = 0.5),
     ## axes
     axis.title.x = element_text(vjust = -1)) +
  ## change legend dot size and title
  guides(colour = guide_legend(override.aes = list(size = 3), 
                               title = "Food Group:"))
```
**Figure 1.** PCA biplot for the variance in macronutrients of raw vegetables (green) and fruits (purple) as reported by the USDA. 

```{r}
### All info biplot
autoplot(fruit_veg_all_pca,
     	data = fruit_veg_all,
     	loadings = TRUE,
     	colour = 'food_group',
     	size = 1.8,
     	alpha = .6,
     	loadings.label = TRUE,
     	loadings.colour = "black",
     	loadings.label.colour = "black",
     	loadings.label.vjust = -0.5) +
  ## dot colors and legend labels
  scale_colour_manual(values = c('seagreen3', 'mediumorchid1'), 
                      labels = c('Vegetables', 'Fruits')) +
  theme_minimal() +
  theme(
     ## legend  
     legend.position = c(.8, .8),
     legend.title = element_text(face = 'bold', size = 10),
     legend.box.background = element_rect(color = 'black', size = 0.5),
     ## axes
     axis.title.x = element_text(vjust = -1)) +
  ## change legend dot size and title
  guides(colour = guide_legend(override.aes = list(size = 3), 
                               title = "Food Group:"))
```
**Figure 2.** PCA biplot of raw vegetable and fruit nutrition information. Points indicating vegetable information are displayed in green, while fruit data points are displayed in purple. 

# Summary

Vector arrow lengths correspond with the eigenvalues for nutrients variables and variance in a given principal component (PC) direction. The angle between arrows indicates correlation. When evaluating only the macronutrients, Figure 1 shows that the variance in food kCal and fiber are mostly explain by PC1, while variance in protein, fat, and sugar are mostly captured by PC2. Additionally, the amount of kCal and fiber in a fruit or vegetable are almost perfectly correlated. Unsurprisingly, the amount of fat and carbohydrates the item has are also correlated to the food's kCals. Protein and sugar as well as fat and sugar, however, are nearly 90$^\circ$ apart, indicating no correlation. Finally, Figure 1 indicates that, on average, raw fruits contain more sugar while raw vegetables contain protein.

When a PCA evaluates these macronutrients as well as the amount of vitamins and minerals, these trends largely persist (Figure 2). This biplot provides further information, such as vegetables contain larger amounts of vitamins and minerals, and vitamin B12 is negatively correlated with sugar. In this new PCA analysis, PC1 largely explains variance in vitamins and minerals, while PC2 is more aligned with macronutrients. Overall, these PCs capture less of the explained variance than those in the first PCA. 


